Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2024-10-16 analysis/paper/article.qmd
working on requirements

```{r}
# stop rendering at this point because the code below is not working
knitr::knit_exit()
```

@@

```{r}
#| label: fig-requirements-over-time

@@

jobdata_requirements <- 
jobdata %>% 
  select(year_ad_posted,
         starts_with("documents_requested")) %>% 
  pivot_longer(-year_ad_posted) %>% 
  mutate(value = case_when(
    value == "Not requested in the job ad" ~ 0,
    value == "One" ~ 1,
    value == "Two (e.g. two syllabi)" ~ 2,
    value == "Three" ~ 3,
    .default = 0
  ))  %>% 
  # trim names a bit
  mutate(name = str_remove(name, "documents_requested_")) %>% 
  mutate(name = str_replace_all(name, "_", " ")) %>% 
  filter(name %in% intresting_requirements) %>% 
  mutate(name = str_wrap(name, 10),
         year_ad_posted = str_replace(year_ad_posted, "-", "\n"))

jobdata_requirements_means <- 
jobdata_requirements %>% # average number requested per year
  group_by(year_ad_posted, 
           name) %>% 
  summarise(mean_n = mean(value))

@@

ggplot(jobdata_requirements_means) +
  aes(year_ad_posted, 
      mean_n,
      group = name) +
  geom_smooth(linewidth = 2,
              colour = "black") +
  geom_jitter(data = jobdata_requirements,
             aes(year_ad_posted, 
                 value),
             alpha = 0.1,
             height = 0.2,
             width =  0.1) +
  facet_wrap(~name,
             scales = "free_y",
             nrow = 1) +
  xlab("Year") +
  ylab("Number requested in job ad") +
  scale_y_continuous(breaks = integer_breaks()) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(size = 8),
        strip.text = element_text( size = 20))

ggsave(here("analysis",
            "figures", 
            "fig-requirements-per-year.png"),
       bg ="white",
       h = 10, # experiment with h and w to get the right size and proportion 
       w = 20,
       units = "in",
       dpi = 900) # make the image nice and crisp

@@

```{r}
#| label: fig-requirements-by-position

# do the requirements differ for associate positions 
jobdata_requirements_by_rank <- 
jobdata %>% 
  mutate(position_title = case_when(
    str_detect(title_of_position_tenure_track_jobs_only, 
               "associate") ~ "associate",
    str_detect(title_of_position_tenure_track_jobs_only, 
               "assistant") ~ "assistant",
    str_detect(title_of_position_tenure_track_jobs_only, 
               "full") ~ "full"))  %>% 
  select(position_title,
         starts_with("documents_requested")) %>% 
  pivot_longer(-position_title) %>% 
  mutate(value = case_when(
    value == "Not requested in the job ad" ~ 0,
    value == "One" ~ 1,
    value == "Two (e.g. two syllabi)" ~ 2,
    value == "Three" ~ 3,
    .default = 0
  )) %>% 
  filter(!is.na(position_title))

@@

ggplot() +
    geom_jitter(data = jobdata_requirements_by_rank,
                  aes(position_title, 
                      value),
                height = 0.05,
                alpha = 0.1) +
    geom_point(data = jobdata_requirements_by_rank_means,
               aes(position_title,
                   mean),
               size = 4,
               colour = "red") +
  facet_wrap( ~ name,
              scales = "free_y") +
  theme_minimal()
```

# topical focus by year

library(googlesheets4)
library(stringi)

topical_foci <-
  read_sheet("https://docs.google.com/spreadsheets/d/1AHq49pIyChcgJ7rawe6KMWkdIBXydCamvg8Jslob8Ec/edit#gid=0",
             sheet = "topic")

jobdata_topic_year <-
  jobdata %>%
  bind_cols(jobdata_topic) %>%
  select(year_ad_posted,
         topical_foci$Category) %>%
  pivot_longer(-year_ad_posted) %>%
  drop_na()

# how many times each topic mentioned?
jobdata_topic_year %>%
  group_by(name) %>%
  summarise(n = sum(value)) %>%
  arrange(desc(n)) %>%
  ggplot() +
  aes(reorder(name, n),
      n)+
  geom_col() +
  xlab("") +
  theme_minimal() +
  coord_flip()

# explore trends over time. put a point on the max year
jobdata_topic_year_tally <-
  jobdata_topic_year %>%
  # exclude those with <20 ads
  filter(!name %in% c("Digital Archaeology",
                      "Pleistocene archaeology",
                      "Mesoamerican Archaeology",
                      "Biological anthropology",
                      "Archaeological theory",
                      "Evolutionary anthropology",
                      "North American archaeology"
  )) %>%
  group_by(year_ad_posted,
           name) %>%
  summarise(n = sum(value)) %>%
  mutate(prop = n / sum(n))

ggplot() +
  geom_smooth(data =  jobdata_topic_year_tally,
            aes(year_ad_posted,
                prop,
                group = name,
                colour = name),
            size = 3,
           span = 0.7,
            se = FALSE) +
  xlab("Year") +
  ylab("Proportion of all ads") +
  scale_colour_brewer(palette = "Dark2") +
  guides(colour = guide_legend("Topic\nfocus",
                               label.position = "bottom")) +
    theme_minimal( base_size = 28) +
  theme(legend.position="bottom")

ggsave(here("analysis",
            "figures", 
            "fig-topic-focus-by-year.png"),
       bg ="white",
       h = 10, # experiment with h and w to get the right size and proportion
       w = 20,
       units = "in",
       dpi = 900) # make the image nice and crisp))

library(googlesheets4)
library(stringi)

method_foci <-
  read_sheet("https://docs.google.com/spreadsheets/d/1AHq49pIyChcgJ7rawe6KMWkdIBXydCamvg8Jslob8Ec/edit#gid=0",
             sheet = "method")

jobdata_method_year <-
  jobdata %>%
  bind_cols(jobdata_method) %>%
  select(year_ad_posted,
         method_foci$Category) %>%
  pivot_longer(-year_ad_posted) %>%
  drop_na()

# how many times each topic mentioned?
jobdata_method_year %>%
  group_by(name) %>%
  summarise(n = sum(value)) %>%
  arrange(desc(n)) %>%
  ggplot() +
  aes(reorder(name, n),
      n)+
  geom_col() +
  xlab("") +
  theme_minimal() +
  coord_flip()

# explore trends over time. put a point on the max year
jobdata_method_year_tally <-
  jobdata_method_year %>%
  # exclude those with <20 ads
  filter(!name %in% c("Material culture analysis",
                      "Ceramic analysis"
  )) %>%
  group_by(year_ad_posted,
           name) %>%
  summarise(n = sum(value)) %>%
  mutate(prop = n / sum(n))

ggplot() +
  geom_smooth(data =  jobdata_method_year_tally,
            aes(year_ad_posted,
                prop,
                group = name,
                colour = name),
            size = 3,
           span = 0.7,
            se = FALSE) +
  xlab("Year") +
  ylab("Proportion of all ads") +
  guides(colour = guide_legend("Method\nfocus",
                               label.position = "bottom")) +
    theme_minimal( base_size = 28) +
  scale_colour_brewer(palette = "Dark2") +
  theme(legend.position="bottom")

ggsave(here("analysis",
            "figures", 
            "fig-method-focus-by-year.png"),
       bg ="white",
       h = 10, # experiment with h and w to get the right size and proportion
       w = 20,
       units = "in",
       dpi = 900) # make the image nice and crisp))
```

```{r}
# EEO
library(stringr)
library(dplyr)
library(stopwords)
library(tidyverse)
library(tidytext)
library(tm)

stopwords <- stopwords("en")

EEO <- jobdata %>%
  select(equal_employment_opportunity_statement, year_ad_posted) %>% 
  drop_na

remove_stopwords <- function(text, stopwords) {
  pattern <- paste0("\\b(", paste(stopwords, collapse = "|"), ")\\b")
  str_remove_all(text, pattern)
}

EEO_clean <- EEO %>%
  mutate(equal_employment_opportunity_statement = str_to_lower(equal_employment_opportunity_statement), 
         equal_employment_opportunity_statement = str_remove_all(equal_employment_opportunity_statement, "[[:punct:]]"),
         equal_employment_opportunity_statement = remove_stopwords(equal_employment_opportunity_statement, stopwords))

#word count for EEO statement
EEO_wordcount <- EEO %>%
  mutate(word_count = str_count(equal_employment_opportunity_statement, "\\S+")) %>%
  print()

jobdata_EEO_year_tally <-
  EEO_wordcount %>%
  group_by(year_ad_posted) %>%
  summarise(n = mean(word_count))

fig_ave_length_EEO_per_year <- 
ggplot(jobdata_EEO_year_tally) +
  aes(year_ad_posted,n) +
  geom_col() +
  ylab("Average length of EEO statement by year") +
  theme_minimal(base_size = 28) +
  guides(x = "none")

ggplot(EEO_wordcount) +
  aes(year_ad_posted, word_count) +
  geom_boxplot() +
  geom_quasirandom()

ggplot(jobdata_requirements_means %>% 
         filter(name == "diversity\nstatement") ) +
  aes(year_ad_posted, 
      mean_n) +
  geom_col()

# here's the list of words of interest
community_word <- c("women","color","religion","race","age","veteran","disability","citizenship","national origin","sex","gender","marital status","genetic information","breastfeeding","sexual orientation","transgender status","gender expression","gender identity","ancestry","income assignment for child support","arrest and court record","domestic or sexual violence victim status","national guard absence","visible minorities","aboriginal persons","indigenous","LGBTQ2S+","pregnancy","childbirth","faith","genetic or family medical history","medical condition","nationality","ethnicity","parental status","accent","ethnic","creed","political affiliation","immigration status")

EEO_word <- EEO_clean %>%
   unnest_tokens(word, 
                 equal_employment_opportunity_statement) %>%
  anti_join(get_stopwords()) %>%
  count(word, sort = TRUE) %>%
  drop_na()

# prepare a dataframe to store the count of words of interest
df <- data.frame(matrix(ncol = length(community_word), 
                        nrow = 0))
colnames(df) <- community_word

for(i in 1: nrow(EEO_clean)){
  df[i, ] <- str_detect(EEO_clean[i, ], 
                   community_word)
}

# detect if the words of interest is present or not.
EEO_clean_community_word <- 
  bind_cols(EEO_clean, df) %>% 
  rowwise() %>% 
  mutate(community_word_tally = sum(c_across(all_of(community_word)), 
                                    na.rm = T)) %>% 
  bind_cols(EEO_wordcount) %>% 
  mutate(eeo_words_total_ratio = community_word_tally / word_count)

ggplot(EEO_clean_community_word) +
  aes(community_word_tally,
      word_count,
      size = eeo_words_total_ratio) +
  geom_point()

ggplot(EEO_clean_community_word) +
  aes(community_word_tally) +
  geom_histogram()
```

```{r}
ggplot(EEO_word %>% 
         filter(n > 20), 
       aes(x = reorder(word, n), 
           y = n)) +
  geom_col() +
  labs(x = "Words", 
       y = "Frequency", 
       title = "Word Frequency") +
  coord_flip()
```

```{r}
# explore QS rankings

qs_rankings <- 
  readxl::read_excel("analysis/data/raw_data/2024 QS WUR by Subject - Public Results (for qs.com)_48.xlsx", sheet = "Archaeology", skip = 12)

jobdata %>% 
  mutate(Institution = str_remove(name_of_hiring_university,
                                  " \\(.*\\)")) %>% 
  mutate(Institution = case_when(
    # modify the school name to match with the name in the QS list
    Institution == "Ohio State University-Main Campus" ~ "The Ohio State University",
  "Cambridge" ~ "University of Cambridge",
  "Cambridge University" ~ "University of Cambridge",
  "Keble College" ~ "University of Oxford",
  "University of California-Berkeley" ~ "University of California, Berkeley (UCB)",
"University of California-Los Angeles" ~ "University of California, Los Angeles (UCLA)",
"University of Tuebingen" ~ "Eberhard Karls Universität Tübingen",
"Universität Tübingen" ~ "Eberhard Karls Universität Tübingen",
"Arizona State University Campus Immersion" ~ "Arizona State University",
"Kiel University" ~ "Christian-Albrechts-University zu Kiel",
"University of Arizona" ~ "The University of Arizona",
"The Pennsylvania State University" ~ "Pennsylvania State University",
"University of Bologna" ~ "Alma Mater Studiorum - University of Bologna",
"University of Sydney" ~ "The University of Sydney",
"University of Copenhagen, Saxo Institute" ~ "University of Copenhagen",
"The University of Texas at Austin" ~ "University of Texas at Austin",

    .default = as.character(Institution)
  )) %>% 
  select(Institution) %>% 
  left_join(qs_rankings) %>% 
  drop_na(`2024`) %>% 
  distinct(Institution, .keep_all = TRUE) %>% 
  View()

>>>>

@@

```{r}
#| label: requirements-over-time

@@

jobdata_requirements <- 
jobdata %>% 
  select(year_ad_posted,
         starts_with("documents_requested")) %>% 
  pivot_longer(-year_ad_posted) %>% 
  mutate(value = case_when(
    value == "Not requested in the job ad" ~ 0,
    value == "One" ~ 1,
    value == "Two (e.g. two syllabi)" ~ 2,
    value == "Three" ~ 3,
    .default = 0
  ))  %>% 
  # trim names a bit
  mutate(name = str_remove(name, "documents_requested_")) %>% 
  mutate(name = str_replace_all(name, "_", " ")) %>% 
  filter(name %in% intresting_requirements) %>% 
  mutate(name = str_wrap(name, 30),
         year_ad_posted = str_replace(year_ad_posted, "-", "\n")) %>% 
  # add line break for prettier plotting
  mutate(year_ad_posted_break = str_replace(year_ad_posted, "-", "-\n"))

jobdata_requirements_means <- 
jobdata_requirements %>% # average number requested per year
  group_by(year_ad_posted, 
           name) %>% 
  summarise(mean_n = mean(value)) %>% 
  # add line break for prettier plotting
  mutate(year_ad_posted_break = str_replace(year_ad_posted, "-", "-\n"))

@@

p_req <- 
ggplot(jobdata_requirements_means) +
  aes(year_ad_posted_break, 
      mean_n,
      group = name) +
  geom_smooth(linewidth = 2,
              colour = "black") +
  geom_jitter(data = jobdata_requirements,
             aes(year_ad_posted_break, 
                 value),
             alpha = 0.1,
             height = 0.2,
             width =  0.1) +
  facet_wrap(~name,
             scales = "free_y",
             nrow = 2) +
  xlab("") +
  ylab("Number requested in job ad") +
  scale_y_continuous(breaks = integer_breaks()) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(size = 8),
        strip.text = element_text( size = 16))

ggsave(filename = here("analysis",
            "figures", 
            "fig-requirements-per-year.png"),
       plot = p_req,
       bg ="white",
       h = 6, # experiment with h and w to get the right size and proportion 
       w = 12,
       units = "in",
       dpi = 900) # make the image nice and crisp

@@

```{r}
#| label: fig-requirements-over-time
#| fig-cap: "Changing requirements in job ads over time. Each point represents an individual job ad. The points have been jittered to avoid overlap. The black line is a locally weighted regression line to identify temporal trends. The grey region is the 95% confidence interval for the regression line."

# show figure from PNG file
knitr::include_graphics(here("analysis",
            "figures",
            "fig-requirements-per-year.png"))
```

Over our ten year study period there have been substantial changes in the instructions to applicants in terms of the type and number of documents that are requested by the search committees [@fig-requirements-over-time]. A cover letter and CV are less requested in more recent years, especially after 2018. The requirement for a diversity statement is rare until 2019-2020, peaks around 2020-2021, then decreases towards the present. Requests for names of recommenders (either zero or three names, rarely only two names) reaches a maximum during 2019-2020 then decreases towards the present. The requirement for a research statement and teaching statement increases after 2015-2016, and becomes more frequent in job ads in more recent years. Requests for course descriptions, syllabus samples, teaching evaluations, transcripts and writing samples are consistently low over time (not shown here).

```{r}
#| label: requirements-by-position

# do the requirements differ for associate positions 
jobdata_requirements_by_rank <- 
jobdata %>% 
  mutate(position_title = case_when(
    str_detect(title_of_position_tenure_track_jobs_only, 
               "associate") ~ "associate",
    str_detect(title_of_position_tenure_track_jobs_only, 
               "assistant") ~ "assistant",
    str_detect(title_of_position_tenure_track_jobs_only, 
               "full") ~ "full"))  %>% 
  select(position_title,
         starts_with("documents_requested")) %>% 
  pivot_longer(-position_title) %>% 
  mutate(value = case_when(
    value == "Not requested in the job ad" ~ 0,
    value == "One" ~ 1,
    value == "Two (e.g. two syllabi)" ~ 2,
    value == "Three" ~ 3,
    .default = 0
  )) %>% 
  filter(!is.na(position_title))   %>% 
  mutate(position_title = str_to_title(position_title)) %>% 
    # trim names a bit
  mutate(name = str_remove(name, "documents_requested_")) %>% 
  mutate(name = str_replace_all(name, "_", " "))

@@

p_req_rank <- 
ggplot() +
    geom_jitter(data = jobdata_requirements_by_rank,
                  aes(position_title, 
                      value),
                height = 0.05,
                alpha = 0.1) +
    geom_point(data = jobdata_requirements_by_rank_means,
               aes(position_title,
                   mean),
               size = 4,
               colour = "red") +
  facet_wrap( ~ name,
              scales = "free_y") +
   scale_y_continuous(breaks = integer_breaks()) +
  xlab("") +
  ylab("Number requested in job ad") +
  theme_minimal()

ggsave(filename = here("analysis",
            "figures", 
            "fig-requirements-by-rank.png"),
       plot = p_req_rank,
       bg ="white",
       h = 6, # experiment with h and w to get the right size and proportion 
       w = 8,
       units = "in",
       dpi = 900) # make the image nice and crisp
```

```{r}
#| label: fig-requirements-by-position
#| fig-cap: "Variation in application requirements by academic position. Each point represents an individual job ad. The points have been jittered to avoid overlap. The red point is the mean value for each group."

# show figure from PNG file
knitr::include_graphics(here("analysis",
            "figures",
            "fig-requirements-by-rank.png"))
```

@fig-requirements-by-position
